Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-21 manuscript/sect_3.tex
section 3

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{dsfont}

\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}

@@

%  Here, place your title and author information.  Note that in 
%  use of the \author command, you create your own footnotes.  Follow
%  the examples below in creating your author and affiliation information.
%  Also consult a recent issue of the journal for examples of formatting.

\title[]{Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories}

%  Here are examples of different configurations of author/affiliation
%  displays.  According to the Biometrics style, in some instances,
%  the convention is to have superscript *, **, etc footnotes to indicate 
%  which of multiple email addresses belong to which author.  In this case,
%  use the \email{ } command to produce the emails in the display.

%  In other cases, such as a single author or two authors from 
%  different institutions, there should be no footnoting.  Here, use
%  the \emailx{ } command instead.

%  The examples below corrspond to almost every possible configuration
%  of authors and may be used as a guide.  For other configurations, consult
%  a recent issue of the the journal.

%  Single author -- USE \emailx{ } here so that no asterisk footnoting
%  for the email address will be produced.

%\author{John Author\emailx{email@address.edu} \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K.}

%  Two authors from the same institution, with both emails -- use
%  \email{ } here to produce the asterisk footnoting for each email address

%\author{John Author$^{*}$\email{author@address.edu} and
%Kathy Authoress$^{**}$\email{email2@address.edu} \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K.}

%  Exactly two authors from different institutions, with both emails  
%  USE \emailx{ } here so that no asterisk footnoting for the email address
%  is produced.

\author
{Carter Allen\emailx{allecart@musc.edu} \\
Department of Public Health Sciences, Medical University of South Carolina, Charleston, SC, U.S.A
\and
Brian Neelon, PhD \\
Department of Public Health Sciences, Medical University of South Carolina, Charleston, SC, U.S.A
\and
Sara Benjamin-Neelon, PhD, MPH, RD \\
Department of Health, Behavior and Society, Johns Hopkins University, Baltimore, MD, U.S.A}

%  Three or more authors from same institution with all emails displayed
%  and footnoted using asterisks -- use \email{ }

%\author{John Author$^*$\email{author@address.edu}, 
%Jane Author$^{**}$\email{jane@address.edu}, and 
%Dick Author$^{***}$\email{dick@address.edu} \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K}

%  Three or more authors from same institution with one corresponding email
%  displayed

%\author{John Author$^*$\email{author@address.edu}, 
%Jane Author, and Dick Author \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K}

%  Three or more authors, with at least two different institutions,
%  more than one email displayed

%\author{John Author$^{1,*}$\email{author@address.edu}, 
%Kathy Author$^{2,**}$\email{anotherauthor@address.edu}, and 
%Wilma Flinstone$^{3,***}$\email{wilma@bedrock.edu} \\
%$^{1}$Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K \\
%$^{2}$Department of Biostatistics, University of North Carolina at 
%Chapel Hill, Chapel Hill, North Carolina, U.S.A. \\
%$^{3}$Department of Geology, University of Bedrock, Bedrock, Kansas, U.S.A.}

%  Three or more authors with at least two different institutions and only
%  one email displayed

%\author{John Author$^{1,*}$\email{author@address.edu}, 
%Wilma Flinstone$^{2}$, and Barney Rubble$^{2}$ \\
%$^{1}$Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K \\
%$^{2}$Department of Geology, University of Bedrock, Bedrock, Kansas, U.S.A.}

@@

%  This will produce the submission and review information that appears
%  right after the reference section.  Of course, it will be unknown when
%  you submit your paper, so you can either leave this out or put in 
%  sample dates (these will have no effect on the fate of your paper in the
%  review process!)

\date{{\it Received October} 2007. {\it Revised February} 2008.  {\it
Accepted March} 2008.}

%  These options will count the number of pages and provide volume
%  and date information in the upper left hand corner of the top of the 
%  first page as in published papers.  The \pagerange command will only
%  work if you place the command \label{firstpage} near the beginning
%  of the document and \label{lastpage} at the end of the document, as we
%  have done in this template.

%  Again, putting a volume number and date is for your own amusement and
%  has no bearing on what actually happens to your paper!

\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
\volume{64}
\pubyear{2008}
\artmonth{December}

%  The \doi command is where the DOI for your paper would be placed should it
%  be published.  Again, if you make one up and stick it here, it means 
%  nothing!

\doi{10.1111/j.1541-0420.2005.00454.x}

%  This label and the label ``lastpage'' are used by the \pagerange
%  command above to give the page range for the article.  You may have 
%  to process the document twice to get this to match up with what you 
%  expect.  When using the referee option, this will not count the pages
%  with tables and figures.

\label{firstpage}

%  put the summary for your paper here

\begin{abstract}
In studies of infant motor development, a crucial research goal is to identify latent clusters of infants that experience delayed development, as this is a known risk factor for adverse outcomes later in life. However, there are a number of statistical challenges in modeling infant development: the data are typically skewed, exhibit intermittent missingness, and are highly correlated across the repeated measurements collected during infancy. Using data from the Nurture study, a cohort of over 600 mother-infant pairs followed from pregnancy to 12 months postpartum, we develop a flexible Bayesian latent class model for the analysis infant motor development. Our model has a number of attractive features. First, we adopt the multivariate skew normal distribution with cluster-specific parameters that accommodate the inherent correlation and skewness in the data. Second, we model the cluster membership probabilities using a novel Pólya-Gamma data-augmentation scheme, thereby improving predictions of the cluster membership allocations. Lastly, we impute missing responses under missing at random assumption by drawing from appropriate conditional skew normal distributions. Bayesian inference is achieved through straightforward Gibbs sampling, and can be carried out in available software such as R.  Through simulation studies, we show that the proposed model yields improved inferences over models that ignore skewness. In addition, our imputation method yields improvements compared to conventional missing data methods, including multiple imputation and complete or available case analysis. When applied to Nurture data, we identified two distinct development clusters: one characterized by delayed “U-shaped” development and a higher percentage of male infants and another characterized by more steady development and a lower percentage of males. The clusters also differed in terms of key demographic variables, such as infant race and maternal pre-pregnancy body mass index. These findings can aid investigators in targeting interventions during this critical early-life developmental window.
\end{abstract}

%  Please place your key words in alphabetical order, separated
%  by semicolons, with the first letter of the first word capitalized,
%  and a period at the end of the list.
%

\begin{keywords}
A key word; But another key word; Still another key word; Yet another key word.
\end{keywords}

%  As usual, the \maketitle command creates the title and author/affiliations
%  display

\maketitle

\setcounter{tocdepth}{3}
\tableofcontents

\newpage

%  If you are using the referee option, a new page, numbered page 1, will
%  start after the summary and keywords.  The page numbers thus count the
%  number of pages of your manuscript in the preferred submission style.
%  Remember, ``Normally, regular papers exceeding 25 pages and Reader Reaction 
%  papers exceeding 12 pages in (the preferred style) will be returned to 
%  the authors without review. The page limit includes acknowledgements, 
%  references, and appendices, but not tables and figures. The page count does 
%  not include the title page and abstract. A maximum of six (6) tables or 
%  figures combined is often required.''

%  You may now place the substance of your manuscript here.  Please use
%  the \section, \subsection, etc commands as described in the user guide.
%  Please use \label and \ref commands to cross-reference sections, equations,
%  tables, figures, etc.
%
%  Please DO NOT attempt to reformat the style of equation numbering!
%  For that matter, please do not attempt to redefine anything!

\section{Introduction}
\label{s:intro}

\subsection{Infant Development Clustering}

Heterogeneity of treatment effects (HTE) (Lanza and Rhoades, 2013).

\subsection{Existing Approaches}

Mixtures of multivariate non-symmetric distributions such as the multivariate skew-normal (MSN) distribution allow for the nuances of the marginal density to be captured with a more parsimonious set of mixture components. Mixtures of MSN distributions have been dealt with previously in a Bayesian context (Fr\"{u}hwirth-Schnatter \& Pyne, 2010), however in these models, focus lies primary on marginal density estimation and inference on the mixture components (i.e. clusters) is not discussed. More recently, the mixtures of skew-$t$ factor analysis (MSTFA) model has been proposed for settings in which cluster-specific inference is of primary interest (Lin \textit{et al.} 2018). However, an important feature not included in the MSTFA is the ability to explain individual-level cluster membership as a function of covariates of interest. Additionally, parameter estimation proposed by Lin et al. for the MSTFA relies on a prohibitively complex EM algorithm and does not enjoy the inferential benefits of a Bayesian approach, namely the ability to incorporate prior information into a model and make posterior probability statements. Our proposed model improves on these previous works by estimating parameters in a Bayesian framework as well as including the ability to fit a multinomial logit regression to cluster membership probabilities using a novel application of data augmentation with the P\'olya Gamma distribution.

\textbf{\textit{Put lit review of Bayesian PG multinomial logistic regression here}}

A ubiquitous feature of repeated measures studies is loss of data due to intermittent missingness and attrition. In the Bayesian setting, the standard approach to dealing with missing data is to perform multiple imputation, whereby $m$ imputed data sets are generated from a specified imputation model. After $m$ complete data sets are obtained, parameter estimates are combined across each data set to produce a final set of parameter estimates (Gelman \textit{et al.} 2013). This approach is not only computationally burdensome, requiring storage and analysis of an $m \times n_{rows} \times n_{cols}$ data array in addition to multiplication of total model run time by a factor of $m$, but it has been shown to produce unreliable inferences (Zhou and Reiter, 2010). We instead include an ``online" imputation imputation step in our Gibbs sampling procedure, whereby missing outcomes are updated at each iteration. This approach greatly increases the number of opportunities for exploration of the missing data parameter space.

\newpage

\section{Nurture Study}
\label{s:nurt}
\subsection{Baseline Demographics and Description of Variables}

\subsection{Statistical Challenges}
\subsubsection{Skewness of Bayley score residuals}
\subsubsection{Attrition and Intermittent Missingness}

\newpage

@@

We note that the multivariate skew normal density can be expressed more compactly in terms of the matrix skew normal (MatSN) density (Chen and Gupta 2005). As we will see in Section 3.6, this specification facilitates prior specification by admitting convenient joint prior distributions for the regression coefficients and scale matrices. In particular, we define $\mathbf{Y}_k$ as the ${n_k \times J}$  response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n \mathds{1}_{(z_i = k)}$ is the total number of observations belonging to cluster $k$ ($k = 1,...,K$). 
$$\mathbf{Y}_{k} \sim MatSN_{n_k \times J}(\mathbf{M},\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)$$
$$\text{vec}(\mathbf{Y}) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})$$
$$\text{vec}(\mathbf{M}) = (\boldsymbol\zeta_1^T,...,\boldsymbol\zeta_{n_k}^T)$$
where $\boldsymbol\zeta_i = \mathbf{x}_i^T \boldsymbol\beta_k$ and $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})$. This implies that the $i^{th}$ row of $\mathbf{Y}_k$, a ${n_k \times J}$ matrix of responses for cluster $k$, is $\mathbf{y}_i \sim MSN_J(\boldsymbol\zeta_i, \boldsymbol\alpha_k, \boldsymbol\Omega_k)$. Using the stochastic representation presented in Section 3.2, the conditional distribution of $\mathbf{y}_i | z_i, t_i$ is $N_J (\mathbf{x}_i \boldsymbol\beta_k + \boldsymbol\psi_k t_i,\boldsymbol\Sigma_k)$. This implies that all responses for cluster $k$ conditioned on $\mathbf{t}_{k}$ is distributed as 
$$\mathbf{Y}_k | \mathbf{t}_k \sim MatNorm_{n_k \times J}(\mathbf{M}_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),$$
where $\text{vec}(\mathbf{M}_k)_{n_kJ \times 1} = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ and $\mathbf{X}_k$ is an $n_k \times p$ design matrix.

@@

where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = \mathds{1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$

@@

\begin{algorithm}
\caption{Gibbs Sampler}
\begin{algorithmic}
\small
\linespread{0.5}
    \State $\mathbf{Define} \ n_{iter};\  n_{burn}; \ K; \ \boldsymbol\theta_{init}; \ \boldsymbol\theta_0$
    \State $n_{sim} := n_{iter} - n_{burn}$
    \State $\boldsymbol\theta := \boldsymbol\theta_{init}$
    \For {$\iota = 1,...,n_{sim}$}
        \State \textsc{I. Conditional Imputation}
          \For {$ i = 1,...,n$}
            \State $\mathbf{Draw} \ \mathbf{y}_i^{miss} \ \text{from} \ N_q(\boldsymbol\mu_i^{miss}, \boldsymbol\Sigma_i^{miss})$
            \State $\mathbf{y}_i := \mathbf{y}_i^{miss} \cup \mathbf{y}_i^{obs}$
         \EndFor
        \State \textsc{II. MSN Regression}
          \For {$ k = 1,...,K$}
            \State $n_k := \sum_{i = 1}^n \mathds{1}_{z_i = k}$
            \For {$i_k = 1,...,n_k$}
              \State $\mathbf{Draw}\  t_i \ \text{from} \ N_{[0,\infty)}(a_i,A)$
            \EndFor
            \State $\mathbf{X^*}_k := $ \texttt{cbind}$(\mathbf{X}_k,\mathbf{t}_k)$
            \State $\mathbf{Draw} \ \mathbf{B^*}_k \ \text{from} \ \text{MatNorm}(\mathbf{B}_k,\mathbf{L}_k^{-1},\boldsymbol\Sigma_k)$
            \State $\mathbf{Draw} \ \boldsymbol\Sigma_k \ \text{from} \ \text{InvWish}(\nu_k, \mathbf{V}_k)$
          \EndFor
        \State \textsc{III. Multinomial Logit}
          \For {$i = 1,...,n$}
            \For {$k = 1,...,K$}
              \State $\pi_{ik} := P(z_i = k|\mathbf{w}_i,\boldsymbol\delta_k)$
              \State $p_{ik} := P(\mathbf{y}_i|\boldsymbol\beta_k^{*T} \mathbf{x}^*_i,\boldsymbol\Sigma_k)$
            \EndFor
            \State $\mathbf{p}_{z_i} := \frac{\mathbf{p}_i \circ \boldsymbol\pi_i}{\mathbf{p}_i \cdot \boldsymbol\pi_i}$
            \State $\mathbf{Draw} \ z_i \ \text{from} \ \text{Categorical}(\mathbf{p}_{z_i})$
            \For {$k = 1,...,K-1$}
              \State $\mathbf{Draw} \ \boldsymbol\delta_k \ \text{from} \ N(\mathbf{M},\mathbf{S})$
            \EndFor
          \EndFor
        \State $\boldsymbol\theta := \{\mathbf{B^*}, \boldsymbol\Sigma, \mathbf{Z}, \boldsymbol\delta \}$
        \State $\mathbf{Store} \ \boldsymbol\theta$
	  \EndFor
\end{algorithmic}
\end{algorithm}

@@

\newpage

\section{Simulation Studies}
\label{s:sim}

\subsection{Simulation to Compare to Multivariate Normal}

\begin{table}[t]

\caption{\label{tab:unnamed-chunk-4}Model results for simulated data with n = 1500, k = 4, p = 1, h = 3, v = 1. 5000 iterations were run with a burn in of 1000. Missingness mechanism was MAR and P(miss) = 0}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}{llllllll}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Cluster 1} & \multicolumn{2}{c}{Cluster 2} & \multicolumn{2}{c}{Cluster 3} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8}
Model Component & Parameter & True & Est. (95\% CrI) & True & Est. (95\% CrI) & True & Est. (95\% CrI)\\
\midrule
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em}MVSN & $\beta_{0}$ & -1.35 & -1.59 (-2.3, -0.94) & -0.18 & -0.24 (-0.53, 0.38) & 0.72 & 0.79 (0.52, 1.02)\\
\hspace{1em}Regression & $\beta_{1}$ & -1.19 & -1.35 (-1.99, -0.84) & -0.09 & -0.33 (-0.62, 0.78) & 1.64 & 1.65 (1.33, 1.91)\\
\hspace{1em} & $\beta_{2}$ & -1.65 & -1.81 (-2.66, -1.41) & -0.47 & -0.62 (-0.89, 0.3) & 1.44 & 1.32 (0.97, 1.6)\\
\hspace{1em} & $\beta_{3}$ & -1.75 & -1.89 (-2.52, -1.37) & -0.22 & -0.32 (-0.63, 0.55) & 2.28 & 2.26 (1.96, 2.5)\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em} & $\sigma_{11}$ & 1 & 1.02 (0.78, 1.19) & 1 & 0.98 (0.77, 1.23) & 1 & 1.06 (0.85, 1.29)\\
\hspace{1em} & $\sigma_{12}$ & -0.32 & -0.19 (-0.33, -0.02) & 0.16 & 0.14 (-0.01, 0.4) & 0.72 & 0.82 (0.62, 1.05)\\
\hspace{1em} & $\sigma_{13}$ & -0.65 & -0.55 (-0.68, -0.35) & 0.72 & 0.7 (0.51, 0.94) & 0.14 & 0.27 (0.1, 0.48)\\
\hspace{1em} & $\sigma_{14}$ & -0.44 & -0.33 (-0.46, -0.13) & 0.5 & 0.48 (0.31, 0.72) & -0.01 & -0.02 (-0.16, 0.16)\\
\hspace{1em} & $\sigma_{22}$ & 1 & 0.92 (0.72, 1.06) & 1 & 0.94 (0.72, 1.22) & 1 & 1.11 (0.87, 1.38)\\
\hspace{1em} & $\sigma_{23}$ & 0.56 & 0.49 (0.33, 0.6) & 0.53 & 0.46 (0.29, 0.73) & -0.1 & 0.08 (-0.1, 0.28)\\
\hspace{1em} & $\sigma_{24}$ & 0.98 & 0.9 (0.7, 1.04) & 0.24 & 0.14 (-0.03, 0.41) & 0.19 & 0.17 (0.01, 0.37)\\
\hspace{1em} & $\sigma_{33}$ & 1 & 0.9 (0.66, 1.04) & 1 & 0.92 (0.72, 1.19) & 1 & 1.26 (1.03, 1.52)\\
\hspace{1em} & $\sigma_{34}$ & 0.56 & 0.51 (0.35, 0.62) & 0.86 & 0.79 (0.58, 1.05) & -0.65 & -0.63 (-0.78, -0.45)\\
\hspace{1em} & $\sigma_{44}$ & 1 & 0.93 (0.72, 1.07) & 1 & 0.93 (0.7, 1.2) & 1 & 1.11 (0.87, 1.36)\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em} & $\psi_{1}$ & -0.33 & -0.02 (-0.84, 0.87) & 0.67 & 0.69 (-0.09, 1.01) & -1 & -0.98 (-1.25, -0.67)\\
\hspace{1em} & $\psi_{2}$ & -0.33 & -0.16 (-0.8, 0.62) & 0.67 & 0.81 (-0.6, 1.14) & -1 & -0.98 (-1.28, -0.6)\\
\hspace{1em} & $\psi_{3}$ & -0.33 & -0.15 (-0.65, 0.89) & 0.67 & 0.72 (-0.35, 1.02) & -1 & -0.8 (-1.15, -0.36)\\
\hspace{1em} & $\psi_{4}$ & -0.33 & -0.18 (-0.84, 0.59) & 0.67 & 0.71 (-0.32, 1.06) & -1 & -1 (-1.32, -0.62)\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em}Multinom. & $\delta_{11}$ & -0.84 & -0.78 (-0.96, -0.59) & -0.84 & -0.78 (-0.96, -0.59) & -0.84 & -0.78 (-0.96, -0.59)\\
\hspace{1em} & $\delta_{12}$ & -0.24 & -0.26 (-0.42, -0.1) & -0.24 & -0.26 (-0.42, -0.1) & -0.24 & -0.26 (-0.42, -0.1)\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{ }}\\
\hspace{1em}Clustering & $\pi_l$ & 0.39 & 0.39 (0.38, 0.4) & 0.26 & 0.26 (0.25, 0.27) & 0.34 & 0.35 (0.33, 0.36)\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Simulation to Compare Imputation Methods}

\subsection{Simulation to Assess Sensitivity to Misspecified K}

\newpage

\section{Application}
\label{s:app}

\item Include both time varying and non-time varying covariates for the within cluster covariate set.

\newpage

\section{Discussion}
\label{s:discuss}

\textit{\textbf{Discuss non-ignorable missingness here}}

\item Discuss how we handle non-ignorable missingness

\section{Appendix}

Put your final comments here.

%  The \backmatter command formats the subsequent headings so that they
%  are in the journal style.  Please keep this command in your document
%  in this position, right after the final section of the main part of 
%  the paper and right before the Acknowledgements, Supplementary Materials,
%  and References sections.

\backmatter

%  This section is optional.  Here is where you will want to cite
%  grants, people who helped with the paper, etc.  But keep it short!

\section*{Acknowledgements}

%  If your paper refers to supplementary web material, then you MUST
%  include this section!!  See Instructions for Authors at the journal
%  website http://www.biometrics.tibs.org

\section*{Supplementary Materials}

\newpage

\subsection{Glossary of Notation}

    \item $\mathbf{Y}$: A $n \times J$ matrix containing all multivariate skew-normal  outcomes such that $y_{ij}$ is the $j^{th}$ outcome observed for subject $i$, where $i = 1,...,n$ and $j = 1,...J$.

    \item $\mathbf{X}$: A $n \times P$ matrix containing all multivariate skew-normal regression covariates such that $x_{ip}$ is the $p^{th}$ covariate value for subject $i$, where $i = 1,...,n$ and $p = 1,...P$.

    \item $\mathbf{B}$: A $P \times J$ matrix containing all multivariate skew-normal regression coefficients such that $\mathbf{B} = \left [ \boldsymbol\beta_1,...,\boldsymbol\beta_J \right ]$, where $\beta_{pj}$ is interpreted as the effect of covariate $p$ on outcome $j$ for $p = 1,...,P$ and $j = 1,...,J$.

    \item $\mathbf{E}$: A $n \times J$ matrix of error terms in the multivariate skew-normal regression model component. $\mathbf{E}$ is made up of row vectors $\boldsymbol\epsilon_i = (\epsilon_{i1},...,\epsilon_{iJ})$, where $ \boldsymbol\epsilon_i \stackrel{iid}{\sim} N_J(0, \boldsymbol\Sigma)$ for $i = 1,...,n$.

    \item $\boldsymbol\Sigma$: A $J \times J$ covariance matrix that defines the correlation between the $p$ multivariate normal outcomes.

    \item $\boldsymbol\Omega$: A $J \times J$ covariance scale matrix that defines the correlation between the $p$ multivariate skew-normal outcomes.

    \item $\boldsymbol\psi$: A $J \times 1$ vector containing the skewness parameter for each outcome.

    \item $\boldsymbol\alpha$: A $J \times 1$ vector containing the skewness parameter for each outcome.

    \item $\mathbf{t}$: An $n \times 1$ vector of truncated normal random effects used in the stochastic representation of the multivariate skew-normal distribution. For $i = 1,...,n$, $t_i \stackrel{iid}{\sim}T_{[0,\infty)}(0,1)$

    \item $\mathbf{X}^*$: A $n \times (P + 1)$ matrix constructed by column binding $\mathbf{t}$ to $\mathbf{X}$

    \item $\mathbf{B}^*$: A $(P+1) \times J$ matrix constructed by row binding $\boldsymbol\psi^T$ to $\mathbf{B}$.

\newpage
\subsection{Derivation of Full Conditional Distributions}

\subsubsection{Multivariate Skew-Normal Regression}

Without loss of generality, we derive the full conditional distributions for the multivariate skew-normal regression model component under the assumption that all observations belong to a single cluster. To make the extension to the case where more than one cluster is specified, simply apply these distributional forms to cluster specific parameters and data. Finally, we assume for the moment that we have complete data for all outcomes for each subject. We extend consider the case of missing data in section (INSERT SECTION).

The multivariate skew-normal regression model can be written as follows in matrix form. 
$$\mathbf{Y} = \mathbf{X}\mathbf{B} + \mathbf{t} \boldsymbol\psi^T + \mathbf{E} = \mathbf{X}^* \mathbf{B}^* + \mathbf{E}$$
The matrix $\mathbf{Y}$ is of dimension $n \times J$. For convenience, we define $\mathbf{X}^*$ as a $n \times (P + 1)$ matrix constructed by column binding $\mathbf{t}$ to $\mathbf{X}$, and $\mathbf{B}^*$ as a $(P+1) \times J$ matrix constructed by row binding $\boldsymbol\psi^T$ to $\mathbf{B}$. We assume that $t_i \stackrel{iid}{\sim}T_{[0,\infty})(0,1)$ and that $\mathbf{E}$ is made of row vectors $\boldsymbol\epsilon_i = (\epsilon_{i1},...,\epsilon_{iJ})$ for $i = 1,...,n$, where $ \boldsymbol\epsilon_i \stackrel{iid}{\sim} N_J(0, \boldsymbol\Sigma)$.

The conditional likelihood for this model is given below. 
$$p(\mathbf{Y}|\mathbf{X}^*,\mathbf{B}^*,\boldsymbol\Sigma) \propto |\boldsymbol\Sigma|^{-n/2}\exp \left \{ -\frac{1}{2} \tr(\mathbf{Y} - \mathbf{X}^* \mathbf{B}^*)^T(\mathbf{Y} - \mathbf{X}^* \mathbf{B}^*)\boldsymbol\Sigma^{-1} \right \}$$

We choose conjugate priors for $\mathbf{B}^*$ and $\boldsymbol\Sigma$ as follows. 
$$\boldsymbol\Sigma \sim \text{inverse-Wishart}(\mathbf{V}_0,\nu_0)$$
$$\mathbf{B}^*|\boldsymbol\Sigma \sim MatNorm_{(m+1), p}(\mathbf{B}_0^*,\mathbf{L}_0^{-1},\boldsymbol\Sigma)$$

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

@@

@@

We note that the multivariate skew normal density can be expressed more compactly in terms of the matrix skew normal (MatSN) density (Chen and Gupta 2005). As we will see in Section 3.6, this specification facilitates prior specification by admitting convenient joint prior distributions for the regression coefficients and scale matrices. In particular, we define $\mathbf{Y}_k$ as the ${n_k \times J}$  response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n {1}_{(z_i = k)}$ is the total number of observations belonging to cluster $k$ ($k = 1,...,K$). 
$$\mathbf{Y}_{k} \sim MatSN_{n_k \times J}(\mathbf{M},\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)$$
$$\text{vec}(\mathbf{Y}) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})$$
$$\text{vec}(\mathbf{M}) = (\boldsymbol\zeta_1^T,...,\boldsymbol\zeta_{n_k}^T)$$
where $\boldsymbol\zeta_i = \mathbf{x}_i^T \boldsymbol\beta_k$ and $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})$. This implies that the $i^{th}$ row of $\mathbf{Y}_k$, a ${n_k \times J}$ matrix of responses for cluster $k$, is $\mathbf{y}_i \sim MSN_J(\boldsymbol\zeta_i, \boldsymbol\alpha_k, \boldsymbol\Omega_k)$. Using the stochastic representation presented in Section 3.2, the conditional distribution of $\mathbf{y}_i | z_i, t_i$ is $N_J (\mathbf{x}_i \boldsymbol\beta_k + \boldsymbol\psi_k t_i,\boldsymbol\Sigma_k)$. This implies that all responses for cluster $k$ conditioned on $\mathbf{t}_{k}$ is distributed as 
$$\mathbf{Y}_k | \mathbf{t}_k \sim MatNorm_{n_k \times J}(\mathbf{M}_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),$$
where $\text{vec}(\mathbf{M}_k)_{n_kJ \times 1} = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ and $\mathbf{X}_k$ is an $n_k \times p$ design matrix.

@@

where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = {1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$

@@

\begin{algorithm}
\caption{Gibbs Sampler}
\begin{algorithmic}
\small
\linespread{0.5}
    \State $\mathbf{Define} \ n_{iter};\  n_{burn}; \ K; \ \boldsymbol\theta_{init}; \ \boldsymbol\theta_0$
    \State $n_{sim} := n_{iter} - n_{burn}$
    \State $\boldsymbol\theta := \boldsymbol\theta_{init}$
    \For {$\iota = 1,...,n_{sim}$}
        \State \textsc{I. Conditional Imputation}
          \For {$ i = 1,...,n$}
            \State $\mathbf{Draw} \ \mathbf{y}_i^{miss} \ \text{from} \ N_q(\boldsymbol\mu_i^{miss}, \boldsymbol\Sigma_i^{miss})$
            \State $\mathbf{y}_i := \mathbf{y}_i^{miss} \cup \mathbf{y}_i^{obs}$
         \EndFor
        \State \textsc{II. MSN Regression}
          \For {$ k = 1,...,K$}
            \State $n_k := \sum_{i = 1}^n {1}_{z_i = k}$
            \For {$i_k = 1,...,n_k$}
              \State $\mathbf{Draw}\  t_i \ \text{from} \ N_{[0,\infty)}(a_i,A)$
            \EndFor
            \State $\mathbf{X^*}_k := $ \texttt{cbind}$(\mathbf{X}_k,\mathbf{t}_k)$
            \State $\mathbf{Draw} \ \mathbf{B^*}_k \ \text{from} \ \text{MatNorm}(\mathbf{B}_k,\mathbf{L}_k^{-1},\boldsymbol\Sigma_k)$
            \State $\mathbf{Draw} \ \boldsymbol\Sigma_k \ \text{from} \ \text{InvWish}(\nu_k, \mathbf{V}_k)$
          \EndFor
        \State \textsc{III. Multinomial Logit}
          \For {$i = 1,...,n$}
            \For {$k = 1,...,K$}
              \State $\pi_{ik} := P(z_i = k|\mathbf{w}_i,\boldsymbol\delta_k)$
              \State $p_{ik} := P(\mathbf{y}_i|\boldsymbol\beta_k^{*T} \mathbf{x}^*_i,\boldsymbol\Sigma_k)$
            \EndFor
            \State $\mathbf{p}_{z_i} := \frac{\mathbf{p}_i \circ \boldsymbol\pi_i}{\mathbf{p}_i \cdot \boldsymbol\pi_i}$
            \State $\mathbf{Draw} \ z_i \ \text{from} \ \text{Categorical}(\mathbf{p}_{z_i})$
            \For {$k = 1,...,K-1$}
              \State $\mathbf{Draw} \ \boldsymbol\delta_k \ \text{from} \ N(\mathbf{M},\mathbf{S})$
            \EndFor
          \EndFor
        \State $\boldsymbol\theta := \{\mathbf{B^*}, \boldsymbol\Sigma, \mathbf{Z}, \boldsymbol\delta \}$
        \State $\mathbf{Store} \ \boldsymbol\theta$
	  \EndFor
\end{algorithmic}
\end{algorithm}

@@
